Optimal Pricing and Informal Sharing: \\ Evidence from Piped Water in Manila

### 2020-05-12 phil_paper/paper_booster_pumps_wjv.tex
pressure running

\usepackage[notablist,nofiglist]{endfloat}

@@

\renewcommand{\thetable}{\Roman{table}}
\renewcommand{\thefigure}{\Roman{figure}}

\renewcommand{\thesection}{\Roman{section}} 
\renewcommand{\thesubsection}{\thesection.\Alph{subsection}}
\renewcommand{\thesubsubsection}{\thesubsection.\roman{subsubsection}}

@@

\date{November, 2017}
\begin{document}

@@

	Households often share connections to public utilities with their neighbors in order to split set-up and maintenance costs.  With microdata from 1.5 million water connections in Manila, this paper estimates water demand when connections are shared and solves the regulator's problem: setting prices that maximize consumer surplus while covering production costs.  In practice, cities often set low fixed prices so that more households connect and high marginal prices to cover production costs.  Instead, the optimal policy covers production costs with high fixed prices, which are split between sharing households, and sets low marginal prices to encourage consumption.  Counterfactual exercises find increases in shared connections, total piped water consumption, and social welfare from the optimal policy compared to current prices in Manila.

@@

\section{Introduction}
 % \footnote{\cite{laffont2005regulation} emphasizes high costs of public funds in developing countries (with a deadweight loss of 2.5 for the Philippines) due to limited ability to enter long-term contracts with government, credit constraints, inefficient taxation, and corruption.}
 % To ensure widespread access while covering costs, a

%  For example, when households can easily share connections, public utilities can charge larger upfront fees for the same levels of access as when sharing is .  [UNCLEAR COMPARISON!!]

% Households also share connections to public utilities with their neighbors (piped water, electricity, mobile phones, etc.) especially in developing countries.\footnote{70\% of African water utilities report frequent sharing of water taps between households.  Over 14\% of urban households in Arba-Minch, Ethiopia report sourcing electricity from their neighbors instead of purchasing directly from the utility (\cite{mustefa2016energy}).  In South Africa as of 2012, 11.9\% of the population report sharing their mobile phone with others.  This percentage jumps to 19.9\% among those living in poverty (\cite{calandro2012mobile}).}

% Connection sharing can have important implications for the optimal balance of fixed and marginal prices for these services.

% Understanding how this sharing behavior may affect optimal prices may have large consequences for public policy since governments actively regulate prices charged by public utilities.

% In developing countries, households often 
% Faced with fixed costs of accessing a good or service, consumers are increasingly .  
% In developing countries, these .

%like water and electricity

Faced with high fixed costs of purchasing and maintaining durable goods, consumers often share these goods with each other.  While online platforms have accelerated peer-to-peer rentals of cars, apartments, and storage space (\cite{horton2016owning}), households in developing countries often share connections to public utilities with their neighbors.\footnote{70\% of African water utilities report frequent sharing of water taps between households.  Over 14\% of urban households in Arba-Minch, Ethiopia report sourcing electricity from their neighbors instead of purchasing directly from the utility (\cite{mustefa2016energy}).  In South Africa as of 2012, 11.9\% of the population report sharing their mobile phone with others.  This percentage jumps to 19.9\% among those living in poverty (\cite{calandro2012mobile}).}  Sharing can have large consequences for optimal pricing, particularly for public utilities.  This paper focuses empirically on pricing for piped water when households share connections.  Improving access to piped water can generate substantial health benefits\footnote{\cite{zhang2016long,kosec2014child,gamper2010impact}, and \cite{galiani2005water} focus on health improvements in developing countries while \cite{cutler2005role} look historically at the US.} as well as improvements in household well-being through time savings and reduced intra-household conflict.\footnote{\cite{devoto2012happiness} provide water through a randomized-controlled trial in Morocco finding few health effects but large benefits to household welfare.}  Yet, the share of urban households without piped water has remained stable at 5\% for middle income countries and at 20\% for low income countries (World Bank [2015]).  While external health and other benefits may merit subsidizing piped water, any subsidies to public utilities are often limited by high costs of public funds and risk of regulatory capture in developing countries (\cite{laffont2005regulation}).  In many cases, prices must be set to fully recover costs.

%Instead of piped water, these households often resort to local deepwells or water vendors.  CITE THE ADB HERE!

%Public utility pricing is especially salient in developing countries because prices must often be set to fully cover costs (\cite{laffont2005regulation}) and because prices determine not only monthly consumption, but also which households can access these basic services.  For example, when households cannot share connections, they may prefer relatively low fixed prices (ie. installation and maintenance fees) so that each household can afford a connection and high marginal prices (ie. prices per gallon or watt) to cover production costs.  However, when households can easily share connections, they may prefer splitting high fixed prices with their neighbors and enjoying low marginal prices from shared connections.  
%The optimal pricing policy depends both on consumer demand and on the costs of sharing connections between neighbors.

To ensure access to piped water while covering costs, cities often subsidize upfront connection fees with high marginal prices for monthly usage.\footnote{\cite{mcintosh2003asian} and \cite{komives2006distributional} from the World Bank and Asian Development Bank argue for low connection fees.  Using a randomized controlled trial from Morocco, \cite{devoto2012happiness} confirm that modest assistance with connection fees can lead to increases in new water connections.}  Over 70\% of developing cities increase marginal prices with monthly usage so that large users cross-subsidize access and usage for smaller, poorer users.\footnote{This strategy is also commonly referred to as ``Increasing Block Tariffs'' or  ``Convex Pricing'' (\cite{borenstein2012redistributional}). \cite{hoque2013state} and \cite{boland2000political} provide overviews of the use of increasing marginal prices in the water sector in developing countries.  The statistic comes from author calculations with data kindly provided by \cite{hoque2013state}.}  These pricing approaches have been shown to work well when each water connection serves a single household.  Recent empirical papers in C{\^o}te d'Ivoire and South Africa predict gains in both access and total welfare from increasing marginal prices (\cite{diakite2009proposal,szabo2015value}).  However, many households in developing cities share connections with their neighbors, which can have large implications for water pricing.

Insecure land-tenure, credit-constraints, and administrative costs often make it difficult for disadvantaged households to purchase their own water connections (\cite{devoto2012happiness}).  As a result, households frequently share connections by fetching from neighbors with hoses and buckets like in Tangiers, Morocco (\cite{devoto2012happiness}) and drawing from single taps that serve many dwellings like in Kumasi, Ghana (\cite{whittington1992possible}).\footnote{\cite{nauges2006water} finds further evidence of widespread water sharing in Sri Lanka.}  While documenting the prevalence of these  sharing networks, research has not yet examined the importance of water sharing for optimal pricing policy.

% often increase the costs of purchasing a water connection particularly for disadvantaged households
% Also, many cities maintain high upfront fees despite policy recommendations otherwise (\cite{hoque2013state}).

This paper integrates sharing behavior into water demand in order to analyze the welfare effects of pricing policies.  I build a model of household water demand that includes both (1) the upfront choice of whether to purchase a connection, share with a neighbor, or use non-piped sources as well as (2) monthly consumption choices.  I also collect new survey evidence on sharing contracts to inform several features of the model design.  The structural estimation leverages administrative data from 1.5 million water connections in Manila from 2010 to 2015, which link to additional survey data on households sharing connections.

These estimates allow me to solve a standard regulator pricing problem: setting prices to cover fixed and marginal costs while maximizing consumer surplus.  I also let the regulator place extra weight on surplus for smaller, poorer users.  In the presence of sharing, the standard approach --- low fixed fees and high marginal prices --- excludes many small-users from piped water by reducing incentives to share connections.  Instead, these households substitute toward non-piped water sources, which have relatively lower marginal prices, zero upfront costs, and varying water quality.\footnote{Common non-piped water sources in Manila include deepwells, tanker trucks, and water refilling stations.}  The opposite approach --- high fixed fees and low marginal prices --- maximizes both access to piped water and welfare, even for smaller users.  High fixed fees are paid in full by households who connect individually and are split between sharing households.  Informal sharing arrangements enable small-users to access piped water from their neighbors for relatively lower marginal prices than those offered by non-piped sources.  By reducing the total number of connections, sharing arrangements also let the city save on installation and maintenance costs, providing an efficient means for extending piped water to low-demand households.\footnote{In this regulatory arrangement, cost savings are passed on to consumers through lower prices.}  
%  Without fixed prices, land-tenure requirements, or administrative expenses, these sources often have low costs of access.

The welfare impacts of these pricing policies depend not only on household-level demand, but also on the relative costs of sharing water; these may include both marginal hassle costs (ie. time spent fetching from neighbors or reduced pressure from re-piping water) as well as savings from splitting the fixed costs of a connection.  Measuring these features presents two central challenges: how can we recover household-level demand when we observe water consumption for connections serving multiple households, and how can we separately identify the costs of sharing and preferences for water?  The empirical approach addresses these challenges in three steps.

The first step builds a tractable model of water sharing to estimate household-level demand from monthly usage data for each connection.  Informed by new survey evidence, this model embeds how (1) households face challenges in observing each other's monthly demand and (2) households often split the monthly water bill evenly.  These assumptions on sharing behavior ensure that connection-level demand can be expressed as a simple function of underlying household demands even in the presence of non-linear prices.  I use variation in the non-linear price schedule (both across and within connections) to identify water demand for each connection before disaggregating preferences to the household-level.  This approach extends previous models of optimization over non-linear prices by including sharing and allowing households to anticipate unobserved shocks to their demand (\cite{moffitt1986econometrics,burtless1978effect}).

To estimate the marginal hassle costs of sharing water, the second step analyzes a quasi-experiment where breaks in service pipes force some households to switch from their own water connections to their neighbors' connections.  I then measure how much less water households use from their neighbors' connections and attribute this gap to the marginal hassle costs of sharing per cubic meter of usage.  This exercise recovers sizable marginal hassle costs per cubic meter on the order of \input{tables/hassle_percentage}\unskip\% of the average marginal water price.\footnote{This revealed-preference approach builds on a descriptive literature that identifies potential sources of water sharing frictions (\cite{whittington1992possible,nauges2006water}).}

The third step maps the share of households using from each source --- connecting alone, sharing with a neighbor, or using from a non-piped source --- into an estimate of each source's relative fixed cost to consumers.  The empirical approach leverages the assumption that given water preferences (from step one) and marginal sharing costs (from step two), any remaining variation in water source choices must be due to differences in their fixed costs.  Since sharing decisions are strategic in nature, I specify a simple game where households sequentially choose their sources and agree to split any surplus from sharing.  By considering both source and quantity choices, this model expands on the demand estimation literature for public utilities that has so far focused almost exclusively on monthly quantity choices (\cite{diakite2009proposal,mcrae2014infrastructure,olmstead2009reduced,szabo2015value}).

This revealed-preference approach to estimating fixed costs is especially relevant in a developing context because on top of the initial connection and monthly service fees, households often face other fixed costs that are difficult to observe, including permitting fees, plumbing maintenance, as well as land-tenure issues in qualifying for a water connection.  In fact, I estimate that the fixed fees paid to the provider (connection and service fees) account for only around half of the total fixed costs.  This structural approach complements previous experimental evidence of fixed costs by \cite{devoto2012happiness}.

%The third step estimates these fixed costs by simulating the sequential game to match cross-sectional variation in water source choices.

With these estimates, I simulate a series of counterfactual policies.  First, I consider a connection fee subsidy policy consistent with recommendations from the World Bank, NGOs, and other researchers to improve water access for the poor (\cite{komives2005water,jimenez2014factors,mcintosh2003asian}).  I discount the fixed fee by 12\% while increasing the tariff schedule proportionally to ensure that the water provider exactly covers costs.  This policy actually reduces access because households stop sharing connections and substitute towards non-piped sources.  Next, I compute an optimal tariff schedule to maximize consumer surplus while exactly covering production costs.  Compared to the current tariff, I find that a simple, two-part tariff characterized by a fixed fee and low, flat marginal price can produce gains on the order of \input{tables/tpt_improvement_percent_sharing_group}\unskip\% of consumer surplus.\footnote{With quasi-linear utility, consumer surplus for a given tariff is calculated as the difference between household utility under this tariff and household utility from using the non-piped, vendor option.}  These gains are driven both by pricing closer to marginal cost as well as by increasing sharing, which reduces the number of connections installed and maintained by the utility.

Finally, I compute a tariff that maximizes consumer surplus only for the bottom 50\% of users and is intended to capture goals of ensuring minimum usage and access as well as income redistribution.  With sharing, this tariff is ineffective at reallocating surplus from large to small users; however in a counterfactual without sharing, this tariff achieves substantial redistribution.  These findings suggest that informal sharing networks may constrain the extent to which non-linear prices can redistribute consumer surplus.

This paper proceeds by first describing the context of water provision in Manila.  Section~\ref{section:data} describes the data.  Section~\ref{section:patternsofwateruse} presents stylized facts about water usage in Manila.  Section~\ref{section:modelofhouseholdwateruse} introduces a model of household water usage and source choices.  The estimation is divided into three steps in Section~\ref{section:specificationandestimation}.  Section~\ref{section:preferenceestimation} uses monthly quantity choices and price variation to estimate preferences.  Section~\ref{section:hasslecostestimation} analyzes a quasi-experiment to recover an estimate of the marginal costs of sharing with a neighbor.  Given these estimates, Section~\ref{section:fixedcostestimation} recovers the relative fixed costs of owning a connection or using from a vendor.  Section~\ref{section:counterfactualexercises} uses these estimates to examine counterfactual pricing policies.  Section~\ref{section:conclusion} concludes.

\section{Water Provision in Manila}

In 1997, citing poor service quality and low coverage of the public water provider, the Philippines conducted the largest water privatization in the developing world.  Two private companies evenly divided the city into east and west zones as part of a public-private partnership.  Both companies invested heavily in infrastructure, which has resulted in near universal availability of piped water services for Manila.  A government regulator oversees these companies and implements rate-of-return regulation: water tariffs are revised periodically in order to ensure that providers recover all operating costs as well as earn a percentage return on any capital investments.  Rate-of-return regulation is common in developing countries since high costs of public funds as well as high risks of regulatory capture limit funding of utilities through government revenue (\cite{laffont2005regulation}).

This paper focuses on one of the providers which serves 1.5 million connections and around 5 million people.  Figure~\ref{figure:tariff} shows the regulated tariff in Manila.\footnote{Marginal prices continue increasing gradually at 20 m3 intervals above 50 m3, which are not pictured in Figure~\ref{figure:tariff}.  Appendix~\ref{appendix:connectionfeeamortization} includes the full tariff as presented to consumers.}  The increasing marginal price is thought to cross-subsidize consumption among low-demand, poorer households through higher bills from large users.  The regulator assigns zero marginal price to the first 10 cubic meters of monthly usage, which is called the ``lifeline'' rate.  This policy is intended to provide a minimum level of consumption for all households.  This tariff may also perform efficient price discrimination if low-usage households are more price-sensitive.
\begin{figure}
\centering
\caption{Tariff in Manila}\label{figure:tariff}
\begin{center}
\includegraphics[scale=.7]{tables/tariff_structure.pdf}\\
\footnotesize{Figure includes average marginal prices from 2010 to 2015. \\  1 USD $\sim$ 50 PhP}
\end{center}
\end{figure}
% detail more about how to get water, with the fixed fee and conne{}ction fee?!
On top of this tariff schedule, the regulator also charges a fixed fee independent of usage totaling 225 PhP/month.  This fee includes both the monthly service charge and the initial connection fee that I have put into monthly terms for the duration of the water connection.\footnote{Appendix~\ref{appendix:connectionfeeamortization} provides more details about how this monthly fee is calculated and amortized.}  The fixed fee allows the provider to recover any operating costs associated with serving the connection.  This form of increasing tariff and fixed fee combination is implemented in over 70\% of developing cities as well as over 50\% of developed cities (\cite{hoque2013state}).

Faced with this fixed fee as well as other costs associated with getting and maintaining a connection such as additional application costs or land tenure insecurity, I find that over 23\unskip\% of households use a neighbor's tap in Manila.  Sharing a connection with a neighbor mechanically increases the total monthly consumption recorded per connection, often doubling the marginal prices faced by these households compared to using a connection alone.  Although most households use piped water either through their own or neighboring connections, \input{tables/vendor_total}\unskip\% of households remain unconnected to the network and instead fetch water primarily from nearby deepwells while some use water refilling stations or other vendors.

\section{Data}\label{section:data}

%Monthly quantity choices across different households and at different prices reveal information about household demand.  Observing multiple quantities choices for the same household allows for a rich characterization of demand heterogeneity at the household-level.  
Evaluating different pricing policies requires measuring demand at the household level as well as costs associated with providing piped water.  Data was gathered through a partnership with one of the two regulated, private providers in Manila who provided both billing records as well as information about the regulatory structure, pricing, and production costs.  Complete billing records for each water connection including the monthly meter reading, the total bill charged, and the precise location for each meter serve as the primary measure of quantity choices.  These data span 2010 to 2015, during which connections grew from 900,000 to 1.5 million water connections, generating nearly 100 million transaction records.  Connections are divided into residential (90\%), semi-business (4\%), commercial (5\%), and industrial (1\%).  This paper focuses on residential and semi-business connections.\footnote{Semi-business connections are largely composed of households that own small roadside stands.  These connections provide a useful source of price variation which is described in Appendix~\ref{appendix:ratechangeelasticity}.}  The monthly meter reading is used as the measure of consumption for each connection.  Meter readings are rounded to the nearest cubic meter limited to monthly readings below 100 $\text{m}^{3}$ in order to exclude pipe leaks as well as meter reading errors (readings above 100 $\text{m}^{3}$ compose less than 0.7\% of the sample).

Since the object of interest for welfare is household-level demand, the billing records are merged to survey data recording information on households using each connection as well as demographics for the owners of each connection.  To evaluate the performance of the water provider, the regulator contracted with a local university to conduct the Public Assessment of Water Services survey in three waves from 2008 to 2012, surveying around 50,000 water connections.\footnote{Appendix~\ref{appendix:pawssampling} describes the sampling design and tests the extent to which it may be correlated with demographic characteristics.  This exercise finds relatively weak correlations in economic terms with key characteristics.}  The connection survey records the total number of households and people using each connection, providing the primary measure of sharing behavior.  The connection survey also includes basic demographics for the owner of the connection.  The demographics used in the analysis include household size, dwelling type (whether apartment, duplex, or single house), number of employed household members, and whether the head of household is employed in a low-skill profession.
%consumption and water source choices at the household-level.  
%To examine both consumption decisions and water source choices at the household-level, I construct a dataset that merges (1) monthly water usage for around 50,000 connections, (2) demographics and water sharing measures for households using these connections, and (3) a cross-section of all households and their water source choices.  Put together, these data form a representative sample of households in half of Metro Manila.

A full analysis of water usage in Manila also requires measuring households that are unconnected to the bulk water system and instead use water from local vendors.  The 2010 Census of Population and Housing records demographics for households that are using water from alternative sources.  Households are defined as connected to the piped network if they report using a ``faucet community water system'' for their cooking, laundry, or bathing needs.\footnote{The census was designed largely for rural populations so it does not directly ask whether households are connected to a piped water network.}  Alternatively, households are considered to be using from a water vendor if they report any other category, which includes deep wells, peddlers, or other sources.\footnote{Since demographic questions in the census and connection surveys are worded differently, Table~\ref{table:pawstocensus} of Appendix~\ref{appendix:pawstocensus} compares measures across the two data sources for households that own connections finding broadly similar patterns.}  Census data is merged geographically with the connection survey for 599 wards in Manila.

%To construct a profile of water source choices for the full population, I randomly sample from the pool of households using from vendors in 599 wards of the city so that the data form a representative sample when combined with households covered by the connection survey in each small area.
%\footnote{Regions are chosen to overlap with ward administrative/political boundaries while ensuring that I am able to obtain a large enough sample of vendor households for each area to perform the counterfactual analysis in Section~\ref{section:fixedcostestimation}.}

For additional information on water sharing relationships between households, I fielded a small, non-representative survey of 600 households via mobile phone application.  This survey asks about issues associated with using different water sources, basic demographics, as well as how the bill and connection fee are paid for sharing households.  Appendix~\ref{appendix:pollfish} compares this survey to the connection survey finding roughly similar rates of sharing and demographic characteristics.

%includes additional details on this survey and finds roughly similar measures between the connection survey and the mobile survey, suggesting that the mobile survey may provide a reasonable descriptive picture of sharing in Manila.

The water provider also maintains call records of all service complaints, which can be linked to individual connections.  This data is used to measure instances where households face unanticipated shocks to their water source, which are used to identify the marginal costs of using water from a neighbor.  Finally, the water provider tracks data on the marginal cost of pumping an additional cubic meter of water as well as the costs associated with installing new connections.  These data provide inputs into the provider's budget constraint, which are necessary for computing counterfactual pricing regimes.

With information on water source and monthly expenditures as well as household income, the 2011 Community Based Monitoring System data cover over 70\% of households in Pasay City --- a large area with a population of 500,000 in downtown metro Manila.  These data are used to impute income for households outside of the sample according to the following demographic measures: age of household head and indicators for low-skilled employment of household head, residing in a duplex, residing in a house, as well as all possible household sizes and number of employed household members.  The imputation also includes all possible interactions between these measures.

%Supplemental survey data from the Community Based Monitoring Survey in Pasay City --- a large city in the center of Metro Manila --- include household income measures for  that are used to approximate household income using the full set of demographic measures and their interactions.

\section{Patterns of Water Use}\label{section:patternsofwateruse}

The welfare consequences of informal sharing networks depend on both the types of households that choose to share with neighbors as well as the costs of these networks.  To provide some descriptive evidence, Table~\ref{descriptives} merges the connection survey of water connections with billing records associated with each connection.  The columns group connections by the number of households they serve: one, two, or at least three.  Demographics for connection owners indicate that smaller, slightly younger, and poorer households are more likely to share water with neighbors.  Sharing households are over twice as likely to reside in duplexes possibly because transporting water between neighbors requires less effort for duplex residents.

Larger numbers of households sharing a water connection are correlated with more total water usage, but less water per person.\footnote{Total People Served by a connection includes both the owning household and any other households using a connection.}  This decline may have many possible drivers: (1) sharing households face higher prices by locating at a higher point on the tariff schedule (Figure~\ref{figure:tariff}), (2) sharing households may face hassle costs in fetching water from neighboring taps, (3) these price effects may incentivize low-demand households to select into sharing relationships, consistent with the demographic patterns, and (4) low-demand households may have idiosyncratic preferences for sharing relationships.  On average, monthly water bills (inclusive of service fees) take up around \input{tables/incshr}\unskip\% of household income, which does not include other unobserved costs in accessing water.  These findings suggest that policies which affect sharing choices can have significant impacts on water consumption as well as monthly expenditures.

%Demographics for connection owners indicate that smaller and younger households are more likely to share water with neighbors.  Because household size is a strong proxy for water demand, this finding suggests that sharing a connection is negatively correlated with household water demand.  Households with lower imputed income more commonly share water connections.  Sharing households are over twice as likely to reside in duplexes possibly because transporting water between neighbors requires less effort for duplex residents.

\begin{table}
\centering
%\captionsetup{justification=centering}
\begin{threeparttable}
\caption{ Water Connections  by Number of Households Served}\label{descriptives}
\input{tables/new_descriptives}
\begin{tablenotes}
\item \footnotesize{The data include billing records merged with the connection survey.  Income is imputed using location, demographic variables, and their interactions.}
\end{tablenotes}
\end{threeparttable}
\end{table}

%Table~\ref{descriptives} summarizes water use and sharing behavior for these connections, merging information from the connection survey with the billing records.

%Also, the mobile survey indicates that only around 30\% of sharing households would substitute to toward a neighbor's connection if they no longer had access to their primary water source.  For this reason, I treat connections reported to serve more than three households as only serving three households.  
Table~\ref{descriptives} also computes the total number of households using from each source.  This calculation requires the additional assumption that households sharing with one or two other households are uniquely linked to one connection.  This way, the population of households connected to the network can be computed by summing the number of households across all connections.  This assumption is consistent with the presence of fixed costs in setting up a connection or a sharing relationship so that households stay committed to one water source.  To the extent that households are drawing water from multiple connections, this assumption may lead to an overestimation of the population engaged in sharing relationships.  With this assumption, \input{tables/shr_individual}\unskip\% of connected households receive their water from an individual connection, \input{tables/shr_1hh}\unskip\% are involved in a sharing relationship with a single neighbor, and \input{tables/shr_2hh}\unskip\% are in a sharing relationship with 2 or more other households.  These proportions do not take into consideration households that are using from water vendors, which are discussed in more detail below.  Each water connection contributes over four years of billing records on average.

Since the connection survey only includes households connected to piped water, census of population and housing data is required to learn about households using from water vendors.  In this data, \input{tables/vendor_total}\unskip\% of the population report using vendors, which is composed of \input{tables/vendor_well}\unskip\% of households using from a ``deep well'' and \input{tables/vendor_peddler}\unskip\% using from a ``peddler'' while the remaining fall into an ``other'' category.  Anecdotal evidence suggests that groundwater from wells often has very low quality.  Appendix Table~\ref{table:pawstocensus} compares means of demographic variables between connection owners from the connection survey and all households from the census.  Demographic variables are similar between connection owners in the connection survey and the census data.  Households using from vendors have lower imputed incomes, greater probability of low-skill employment, and slightly fewer members than own-use households.  Across most dimensions, households using from vendors more closely resemble households that share with their neighbors as opposed to households that use their connections individually.

%  While demographics are broadly similar between the first and second columns, any differences are likely driven by two factors.  First, connection owners may have different demographic characteristics than households using from a neighbor's tap.  This explanation is consistent with larger household sizes and lower rates of low skilled employment among connection owners in the first column relative to connection owners and other users captured in the second column.  Second, demographic measures may not be exactly comparable across these surveys.  Consistent with this theory, easily measured demographics like age of household head, household size, and total household members employed are very similar across the first and second columns; however, more subjective measures such as level of skill in employment and type of dwelling exhibit greater variation across data sources.  For this reason, the imputation strategy used in Section~\ref{section:fixedcostestimation} also relies on local geographic measures alongside these demographic measures.

\section{A Model of Household Water Use}\label{section:modelofhouseholdwateruse}
% Given that households derive utility from a water source through a stream of water use, 
% Taking these observed choices as inputs, this model is designed to recover both preference parameters for water as well as the shadow costs of using different water sources.

This section presents a stylized model of water source choice and monthly quantity decisions at the household level.  This model allows for an analysis of the impacts of pricing policies on both source and quantity margins, which are often equally relevant to policymakers (\cite{mcintosh2003asian,komives2005water}).  Using both margins also enables the estimation to include cross-sectional variation in sources alongside time-series variation in quantities to recover a broad set of demand and cost parameters.

Let the city of Manila be partitioned into neighborhoods, where each neighborhood is defined by a set of households that are physically close enough to share water connections with each other.  In the context of Manila, these neighborhoods are often characterized by dense clusters of houses or duplexes.  This partition embeds the assumption that neighborhoods are mutually exclusive so that a household located in one neighborhood cannot share water with households in a nearby neighborhood.  Restricting the choice set in this way is necessary for computation but may in some cases overestimate the costs that households face in finding a sharing partner.\footnote{This limitation is discussed in more detail in Section~\ref{section:fixedcostestimation}.}

In the first step, each household makes a one-time decision about their water source between 1) purchasing a connection, 2) using a neighbor's connection, or 3) buying from a local water vendor.  These decisions result in a set of water connections for each neighborhood ranging from an empty set (if all households use from a vendor) to the whole neighborhood (if each household uses an individual connection).  Let each connection, $C$, serve $S$ households that have decided to share the connection.  For ease of notation, let the subscript $C$ indicate connection-level attributes, summing over all households using from connection $C$.  This model does not allow for households to choose different water sources over time, which is consistent with large fixed costs involved in initially purchasing a connection or establishing a sharing relationship.  Section~\ref{section:hasslemodel} extends the model to allow for cases where households may exogenously lose access to their current water source and must substitute toward another nearby source.

In the second step, each household then chooses how much water to use each month from their source.  Since water sources provide utility through monthly consumption, it is useful to describe the decision process in reverse order, first characterizing monthly water consumption choices (taking the initial source choice as given) and second, characterizing the initial water source choices.

\subsection{Consumption Choices for Households Using Piped Water}\label{section:consumptionchoice}
% The goal is to take water consumption data and map it into preferences...
% Water consumption data for each connection aggregates the monthly consumption for all households as well as individuals within these households. 
% Modeling this behavior is not only useful for accurately characterizing demand, but may also play an important role in measuring the potential welfare losses from these non-linear pricing schedules.  
%(CITE JAYACHANDRAN AND JACK)

%This section details how households choose their monthly quantities of consumption holding their water source fixed.

%For households using piped water, they have to consider four aspects when choosing how much to use each month:  (1) households may have difficulty targeting their exact monthly consumption levels since meters can be hard to observe and many family members use the same tap; (2) increasing marginal prices in the water tariff mean that households face non-linear budget constraints; (3) households that share a connection need to determine how to divide the monthly water bill; and (4) households fetching from a neighbor may face additional hassle costs in transporting water.

%Since the data record monthly consumption for each connection, this model first introduces individual household demand before aggregating household demand at the connection-level for estimation.

In choosing how much water to consume each month, households using piped water face the difficult task of coordinating water consumption among all household members.  For example, consider the head of a household trying to manage water consumption across all of their household members each month.  This task would involve monitoring the meter located outside of their dwelling and coordinating water use among all members.\footnote{Given that many developing countries feature steeply increasing prices with monthly usage, these coordination challenges may magnify any distortions caused by non-linear pricing.}  Consistent with these challenges, mobile survey data indicate that \input{tables/no_info}\unskip\% of surveyed household members do not recall how much they paid for water in the last month while \input{tables/no_use}\unskip\% do not recall the amount that their household consumed in the previous month.

As a simple way to capture this coordination task, let each household $i$ choose a goal for water use each month, $w_i$, knowing that unobserved consumption shocks, $\epsilon_i$, may increase or decrease their realized amount of usage.  Consumption shocks, $\epsilon_i$, are assumed to be unobserved to the household and independent across time, bringing the total realized water consumption each month to, $W_i=w_i+\epsilon_i$.  Household utility over total water, $W_i$ and a numeraire good, $x_i$ is assumed to take the following shape:
\begin{align}\label{equation:umaxfirst}
\begin{split}
U(x_i,W_i) &=  x_i + (1-H_i) W_i - \frac{1}{2\alpha_i } ( W_i - \Gamma_i + \alpha_i )^2  \\
W_i &= w_i + \epsilon_i \\
\epsilon_i &\sim \mathcal{N}(0,\Sigma_{\epsilon})
\end{split}
\end{align}
Following from quasi-linear preferences, this quadratic form for utility eliminates income effects and includes a natural satiation point for water consumption, capturing the tendency of households to target specific consumption levels despite low or zero marginal prices as shown in Figure~\ref{figure:consumptionhistogram}.\footnote{This utility form builds on work by \cite{einav2013selection} who use a similar form to characterize extensive and intensive choices of health insurance.}   Household preferences for water are characterized by $\Gamma_i$, which can be interpreted as a fixed preference for water each month and $\alpha_i$, which captures the extent to which households are sensitive to price.  The exogenous shock to consumption, $\epsilon_i$, is assumed to be drawn from a multivariate normal distribution, $\Sigma_{\epsilon}$ with zero mean and variance terms, $\sigma_{\epsilon}^2$, allowing for correlation with neighboring households $i$ and $j$ through term ${\rho}^{\epsilon}_{i,j}$.  Although households cannot anticipate the exact realization of $\epsilon_i$, they are assumed to understand the distribution of possible consumption shocks, allowing them to maximize expected utility.

Households that are using water from a neighbor's tap may face an additional utility cost per gallon of water $H_i$.  This hassle cost captures any inconvenience associated with using an additional unit of water from a neighbor's tap.\footnote{Accordingly, the connection owner is assumed to face zero hassle cost.}  The mobile survey reveals that among households accessing water from a neighbor's connection, \input{tables/access_fetch}\unskip\% fetch water with containers, \input{tables/access_single_tap}\unskip\% access water from a single hose/pipe provided by the connection owner, and \input{tables/access_pipes}\unskip\% are able to fully connect their household plumbing near the water meter.  These different modes of access suggest that sharing households may face higher time costs fetching water or lower water pressure from piping water additional distances.

To account for idiosyncratic preference shocks for water each month, let household's fixed preference, $\Gamma_{i}$, be decomposed into a static, household-specific component, $\gamma_i$, and an idiosyncratic component, $\eta_{i}$, which is assumed to follow a multivariate normal distribution among sharing partners:
\begin{align*}
\Gamma_{i} &= \gamma_i + \eta_{i} \\
\eta_{i} &\sim \mathcal{N} (\textbf{0},\Sigma_{\eta}) 
\end{align*}
This preference shock is also assumed to be independent of the consumption shocks, $\epsilon_C$.  Aggregating these preferences shocks over all households sharing connection, $C$, produces a joint shock, $\eta_C$, according to the convolution of these random normal variables with a variance equal to $\sum_{i=1}^{S} \sigma_{\eta}^2 + \sum_{i=1}^{S} \sum_{j>i}^{S} \rho^{\eta}_{i,j} \sigma_{\eta}^2$.

%To develop notation for sharing, normalize $i=1$ to be the owner of the connection and define the total number of sharing households for a particular connection as $S$.  Therefore, all households, $i\neq1$, using from a neighbor's tap have a positive hassle cost, $H$, while the connection owner, $i=1$, has zero hassle cost.

%FIX HERE!

%Without data on the sharing mode and distance between households for each connection, the model assumes all sharing households face a homogeneous hassle cost, $H_i$, from using a neighbor's connection.\footnote{Section~\ref{section:hasslecostestimation} discusses the extent to which this assumption may bias estimates.  The partitioning of the city into sharing neighborhoods also imposes the implicit assumption that hassle costs rise substantially between neighborhoods so that no household has an incentive to share with households in another neighborhood.}

% , normalized as $i>1$,

\subsubsection{Determining the Budget Constraint}

>>>>

% \usepackage[notablist,nofiglist]{endfloat}

@@

% \renewcommand{\thetable}{\Roman{table}}
% \renewcommand{\thefigure}{\Roman{figure}}

% \renewcommand{\thesection}{\Roman{section}} 
% \renewcommand{\thesubsection}{\thesection.\Alph{subsection}}
% \renewcommand{\thesubsubsection}{\thesubsection.\roman{subsubsection}}

@@

\date{Current Version: March, 2019}
\begin{document}

@@

	% Households often share connections to public utilities with their neighbors in order to split set-up and maintenance costs.  With microdata from 1.5 million water connections in Manila, this paper estimates water demand when connections are shared and solves the regulator's problem: setting prices that maximize consumer surplus while covering production costs.  In practice, cities often set low fixed prices so that more households connect and high marginal prices to cover production costs.  Instead, the optimal policy covers production costs with high fixed prices, which are split between sharing households, and sets low marginal prices to encourage consumption.  Counterfactual exercises find increases in shared connections, total piped water consumption, and social welfare from the optimal policy compared to current prices in Manila.

@@

\section{Introduction}

- Discusses use of booster pumps in Clean Water Act 2004:  https://www.manilatimes.net/2018/11/04/legal-advice/dearpao/use-of-booster-pumps/461829/  http://r12.emb.gov.ph/ra-9275-the-philippine-clean-water-act/

- https://idl-bnc-idrc.dspacedirect.org/bitstream/handle/10625/14878/108376.pdf?sequence=5

- illegal booster pumps :  Consequence of El Niño 1997-98 on Manila water resources: new perspectives in water shortage management

- seizing booster pumps : https://timesofindia.indiatimes.com/city/navi-mumbai/booster-pumps-used-by-some-socs-leading-to-water-crunch/articleshow/68856763.cms , https://timesofindia.indiatimes.com/city/nagpur/215-booster-pumps-seized/articleshow/69804589.cms
